---
title: "GDAT - 515: Assignment 2"
author: "John Viviani"
date: "1/25/2020"
output: 
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
c("dplyr",
  "tidyverse",
  "ggthemes",
  "ggimage", 
  "RCurl",
  "gridExtra"
  ) -> package_names  
for(package_name in package_names) {
  if(!is.element(package_name, installed.packages()[,1])) {
     install.packages(package_name,
                      repos = "http://cran.mtu.edu/")
  }
  library(package_name, character.only=TRUE,
          quietly=TRUE,verbose=FALSE)
}
rm(list=c("package_name", "package_names")) # clean up the environment

options(scipen = 999)
```


## Who Was the Most Improved NFL Team 4 weeks Into the 2019 NFL Season?

The dataset I acquired can be found here: https://github.com/fivethirtyeight/data/tree/master/nfl-elo

It comes from a github repository featuring datasets from fivethirtyeight. The dataset I intend to use contains NFL game-by-game Elo ratings dating all the way back to 1920. I'm aiming to take the difference between each team's Elo rating after their week 4 games and before their week 1 games. This will give a measure of how much each team improved after the first quarter of the season compared to their initial week 1 rating.

The Elo ratings I'm using are based on each team's performance last season. An average team has an Elo rating of 1500. If a team wins, their Elo rating will always improve, and if they lose, their Elo eating will always decline. Various factors are taken into account when assessing how much a team's Elo rating will change (e.g. strength of schedule), and it's meant to be used as a metric to compare each of the team's to one another.

```{r}
# Desktop
read.csv("C:/Users/JVivs/Documents/COLLEGE/GRAD SCHOOL/GDAT515/nfl_elo.csv") -> nfl

# Filtering for weeks 1 through 4 in the 2019 NFL season
nfl[16275:16337,] -> nflteam_4_weeks

str(nflteam_4_weeks)

# Filtering for the variables of interest
# Data is organized by game; therefore, collecting all home team stats in one data frame and all away team stats into another data frame to later combine
nflteam_4_weeks %>%
  select(., "date", "team1", "elo1_pre", "elo1_post") -> df4weeks_team1

nflteam_4_weeks %>%
  select(., "date", "team2", "elo2_pre", "elo2_post") -> df4weeks_team2

# Giving both data frames a common column name to bind later
df4weeks_team1$team <- df4weeks_team1$team1
df4weeks_team2$team <- df4weeks_team2$team2

# Binding the two data frames
newdf_4weeks_team <- bind_rows(df4weeks_team1, df4weeks_team2)

newdf_4weeks_team %>%
  select(., 1,3,4,5,7,8) -> newdf_4weeks_team

# Same thing for pre and post qb values
newdf_4weeks_team$elo1_pre[is.na(newdf_4weeks_team$elo1_pre)] <- newdf_4weeks_team$elo2_pre[is.na(newdf_4weeks_team$elo1_pre)]

newdf_4weeks_team$elo1_post[is.na(newdf_4weeks_team$elo1_post)] <- newdf_4weeks_team$elo2_post[is.na(newdf_4weeks_team$elo1_post)]

# Arranging by date (Start from week 1 to week 4)
newdf_4weeks_team %>%
  arrange(., date) %>% 
  select(., 1:4) %>%
  rename(., team_code = team, elo_pre = elo1_pre, elo_post = elo1_post) -> newerdf_4weeks_team

# Creating a difference column (elo after week 4 - elo before week 1)

# Week 1 data
newerdf_4weeks_team %>%
  slice(., 1:32) %>%
  arrange(., team_code) -> week1_team

# Week 4 data (Including week 3 data for SF and NYJ since they had bye weeks in week 4)
newerdf_4weeks_team %>%
  slice(., 78, 82, 97:126) %>%
  arrange(., team_code) -> week4_team

week4_team$elo_post - week1_team$elo_pre -> week4_team$diff_value
```

I found a dataset that features each of the 32 NFL teams' logos at https://statsbylopez.netlify.com/post/nfl-team-logos-using-ggimage/

To make this bar graph visually appealing, I wanted to add each team's logo to their respective performance measure. In order to do this, I need to merge the dataset created above with this dataset on a common column name (in this case, team_code). Once merged, I can use ggimage to load in the team's logo through the use of the url column in the provided dataset which leads to a PNG of their respective logo.


```{r}
# NFL Team Logos
url.logo <- getURL("https://raw.githubusercontent.com/statsbylopez/BlogPosts/master/nfl_teamlogos.csv")
df.logos <- read.csv(text = url.logo)

str(df.logos)

# Converting logo column to character
as.character(df.logos$team_code) -> df.logos$team_code

# Renaming two team abbreviations to match those in newerdf_4weeks
df.logos$team_code[18] <- c("LAR")
df.logos$team_code[32] <- c("WSH")

# Merging the two data frames together by the team_code column so logos can load in from URL provided in df.logos
merge(week4_team, df.logos, by = "team_code") -> finaldf_team
```

```{r}
# Arranging by highest to lowest
finaldf_team %>%
  arrange(., desc(diff_value)) -> finaldf_team

# Plotting the elo differences
ggplot(finaldf_team, aes(x = reorder(team_code, -diff_value), y = diff_value)) + 
  geom_image(aes(image = url), size = 0.04) +
  geom_bar(aes(fill = diff_value < 0), stat = "identity", alpha = 0.5) + 
  scale_fill_manual(guide = FALSE, breaks = c(TRUE, FALSE), values = c("green3", "red3")) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  xlab("NFL Team") + 
  ylab("Elo Rating") + 
  ggtitle("NFL Team Improvement According to Elo Rating", subtitle = "Through 4 Weeks of the 2019 NFL Season") -> diff_plot_team

diff_plot_team

```

This visualization allows us to see which teams improved from their initial elo ratings (colored green), and which teams regressed (colored red). I arranged it in a descending order so that we can read the graph left to right. This also allows us to quickly answer the question posed as to which team improved the most through 4 weeks of the 2019 NFL Season (San Francisco 49ers). It's important to note that the 49ers and Jets did not play games in week 4 due to having a bye week; so I included their post-week 3 Elo rating into the dataset, and substracted their initial week 1 ratings from that. 

By looking at this graph, we can also see which team regressed the most (Miami Dolphins), so I think it's beneficial having a bar graph in this order as it allows us to glean on other insights aside from the initial question.


